Supplementary Materials: Separating random and deterministic sources of computational noises in explore-exploit decisions

### 2025-04-21 Plos Comp Bio - revision v1/manuscript_DeterministicRandomNoise.tex
u

To gain more quantitative insight into these results, we computed theoretical values for the choice inconsistency for the purely deterministic and purely random noise cases.  For purely deterministic noise this computation is simple because people should make the exact same decisions each time in repeated games, meaning that $p(\mbox{inconsistent}) = 0$ in this case. For purely random noise, the two games should be treated independently, allowing us to compute the choice inconsistency in terms of the probability of choosing the low mean option, $p(\mbox{low mean})$, as
\begin{equation*}
	\begin{split}
		p(\mbox{consistent}) &= p(\mbox{low mean})^2 + p(\mbox{high mean})^2\\
		&= p(\mbox{low mean})^2 + (1-p(\mbox{low mean}))^2\\ 
		\mbox{hence},\quad p(\mbox{inconsistent}) &=  
		1 - p(\mbox{consistent}) = 
		2 p(\mbox{low mean})(1-p(\mbox{low mean}))
	\end{split}
\end{equation*}

@@

For each pair of repeated games, the set of forced-choice trials are exactly the same, so the deterministic noise, $n_{det}$, should be the same while the random noise, $n_{ran}$ may be different. This is exactly how we distinguish deterministic noise from random noise. In symbolic terms, for repeated games $i$ and $j$,  $n_{det}^i=n_{det}^j$  and $n_{ran}^i \neq n_{ran}^j$.

@@

@@

Parameter recover in this task was good for this model, with fit values for $\sigma_{ran}$ and $\sigma_{det}$ showing strong correlations with their simulated values in both horizon (H) conditions (For $\sigma_{ran}$, $R = 0.91 (\mbox{H} = 1) \mbox{ and } 0.84 (\mbox{H} = 6), p < 0.001$, For $\sigma_{det}$, $R = 0.60 (\mbox{H} = 1) \mbox{ and } 0.59 (\mbox{H} = 6), p < 0.001$). However, while the relationship was near perfect for random noise ($\frac{\mbox{\small Recovered } \sigma_{ran}}{\mbox{\small Simulated } \sigma_{ran}} = 1.01$), there was a systematic bias to underestimate the level of deterministic noise by about 32\% ($\frac{\mbox{\small Recovered } \sigma_{det}}{\mbox{\small Simulated } \sigma_{det}} = 0.68$). Despite this underestimation of deterministic noise in both horizon conditions, the difference in deterministic noise between horizons is much better captured (see Supplementary Materials section 2.2). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizon conditions. In addition, we see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. \added[id=siyu]{To ensure that the joint increase of random and deterministic sources of noise is not caused by a limitation of the fitting procedure, we also calculated the correlation between ground-truth values of random noise, and best-fitting values of deterministic noise (and vice versa), and they do not correlate (Supplementary Figure S9).}
	% $\frac{\mbox{\small Recovered } \Delta \sigma_{det}}{\mbox{\small Simulated } \Delta \sigma_{det}} = 0.74$

In addition to the conventional subject-level parameter recovery analysis presented here, we also performed parameter recovery analysis that examined how faithful the full posterior distribution of group-level parameters can be recovered in simulated data (Supplementary Figure S5, S6). Qualitatively, we also showed that our way of modeling deterministic noise is capable of capturing known deterministic processes intentionally omitted from the full model (Supplementary Figure S4). Full details of these additional analysis are presented in the Supplementary Materials.

@@

\subsubsection*{Model-based results} 
Posterior distributions over the group-level means of the deterministic and random noise standard deviation $\sigma_{det}$ and $\sigma_{ran}$ are shown in Figure \ref{fig:mb1} and Supplementary Figure S11. Consistent with our model-free results, we see that both random and deterministic noise are non-zero. Numerically, random noise is about 2-3 times larger than the deterministic noise. By computing the posterior distribution of $\sigma^2_{det}/(\sigma^2_{det}+\sigma^2_{ran})$, our data suggests that 14.25\% of the variability in random exploration is accounted for by deterministic noise ([4.90\%, 28.81\%], 95\% CI). In addition, we find that both random and deterministic noise increase with horizon. This increase was larger for random noise (mean = 7.13, 100\% of samples showed an increase in random noise with horizon) than deterministic noise (mean = 2.59, 98.64\% of samples showed an increase in deterministic noise with horizon). But intriguingly, the relative increase in both types of noise was similar (Figure \ref{fig:ratio}). That is, when we compute the relative increase in deterministic noise with horizon, $\sigma^{det}_{horizon6}/\sigma^{det}_{horizon1}$, it is very similar to the relative increase in random noise with horizon $\sigma^{ran}_{horizon6}/\sigma^{ran}_{horizon1}$. \added[id=siyu]{We also fit alternative models that fit random and deterministic noises separately for [1 3] and [2 2] conditions, and a model with an alternative definition of information bonus $dI$. These results are presented in the supplemental materials (Supplemental Figure. S17, S18). }

@@

As a control, we also applied posterior predictive checks on alternative models that consider only deterministic or only random noise, and these reduced models fail to capture all qualitative patterns (Supplementary Figure S14, S15). Full details of this analysis can be found in Supplementary Materials section 3.2.

@@

Although the random noise interpretation is theoretically appealing, our approach, while an improvement on previous methods, is not without limitations. Most important is that our measure of `random' noise is only an upper bound on the true level of randomness and that, in principle, the random decision noise could be lower. Specifically, in our model, what we labeled random noise was really `non-stimulus-driven variability'. While this non-stimulus-driven variability could be driven by truly random stochastic processes, it could also be driven by deterministic processing that is unrelated to the stimuli in the task. For example, such deterministic noise could be driven by differences in where people look, or for how long they look, or by whether they were fidgeting or scratching their nose \citep{Musall2019}. \added[id=siyu]{Cross-game deterministic strategies and stimuli from previous games (e.g., memory of previously seen game) were also treated as random noise in our model \citep{WYART2016, Wyart_2018}.} In addition to this conceptual limitation in measuring deterministic noise, parameter recovery simulations suggest that our estimation method also slightly underestimates deterministic noise (see Figure \ref{fig:paramrecover_main}, Supplementary Figure S5, S6). As a result, from both a conceptual and methodological perspective, it is possible that the remaining 86\% of the decision noise that is not stimulus-driven noise, could be deterministic.

@@

\added[id=siyu]{We will show that changes in reward processing can be equivalent to changes in noise in decision models (see also \cite{Cinotti19}).}. To show how a change in reward processing could affect random and deterministic noise, consider the simple decision model we introduced in Equation \ref{eq:newmodel}. In this model, choice is determined by the sign of the difference in utility $\Delta Q$ between the two options, where
\begin{equation}
	\label{eq:noise1}
	\Delta Q = 
	\Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Now imagine a case where the reward signal is scaled by a factor $\beta$.  In this case, $\Delta Q$ becomes
\begin{equation}
	\label{eq:noise2}
	\Delta Q = 
	\beta \Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Because the choice only depends on the sign of $\Delta Q$, scaling $\Delta Q$ by a factor of $1/\beta$ will not change the behavior of the model.  Thus, if we divide both sides of the above equation by $\beta$ we get
\begin{equation}
	\label{eq:noise3}
	\Delta Q / \beta = 
	\Delta R + A \Delta I / \beta + b / \beta + n_{det} / \beta + n_{ran} / \beta
\end{equation}
which is equivalent to a scaling of both deterministic and random noise by the same factor $1/\beta$. Thus, one interpretation of our result that both deterministic and random noise change across horizons with the same ratio, is that this reflects a change in reward processing. That is, the reward signal is reduced in the longer horizon condition (smaller $\beta$ in horizon 6 than horizon 1).  %In the alternate interpretation, where deterministic and random noise are varied independently it is not clear why they should change by the same scale factor.

>>>>

To gain more quantitative insight into these results, we computed theoretical values for the choice inconsistency for the purely deterministic and purely random noise cases.  For purely deterministic noise this computation is simple because people should make the exact same decisions each time in repeated games, meaning that $p(\mbox{inconsistent}) = 0$ in this case. For purely random noise, the two games should be treated independently. \added[id=siyu]{Since repeated decisions mean that participants either choose the low-mean option twice, or choose the high-mean option twice, we could predict} the choice inconsistency, $p(\mbox{inconsistent})$, based on the probability of choosing the low mean option, $p(\mbox{low mean})$, as
\begin{equation*}
	\begin{split}
		p(\mbox{consistent}) &= p(\mbox{low mean})^2 + p(\mbox{high mean})^2\\
		&= p(\mbox{low mean})^2 + (1-p(\mbox{low mean}))^2\\ 
		\mbox{hence},\quad p(\mbox{inconsistent}) &=  
		1 - p(\mbox{consistent}) = 
		2 p(\mbox{low mean})(1-p(\mbox{low mean}))
	\end{split}
\end{equation*}

@@

For each pair of repeated games, the set of forced-choice trials are exactly the same, so the deterministic noise, $n_{det}$, should be the same while the random noise, $n_{ran}$ may be different. This is exactly how we distinguish deterministic noise from random noise. In symbolic terms, for repeated games $i$ and $j$,  $n_{det}^i=n_{det}^j$  and $n_{ran}^i \neq n_{ran}^j$. \added[id=siyu]{While an increase in either random or deterministic noises could lead to higher $p(\mbox{low mean})$, an increase in random noise predicts higher $p(\mbox{inconsistent})$ while an increase in deterministic noise predicts lower $p(\mbox{inconsistent})$. }

@@

\added[id=siyu]{We also fit a series of reduced and alternative models to the data. This includes reduced models that assume only deterministic or random noises. We also fit models in which the standard deviation of random and deterministic noises $\sigma_{det}$ and $\sigma_{ran}$ are estimated separately for [1 3] and [2 2] information conditions. Lastly, we fit a model with an alternative definition of $\Delta I$ that $\Delta I$ is defined to be difference between the variances of rewards shown on the forced-choice trials. Results on these model variants are presented in the Supplementary Materials.}

@@

Parameter recover in this task was good for this model (Figure \ref{fig:paramrecover_main}, Supplementary Figure S7, S8), with fit values for $\sigma_{ran}$ and $\sigma_{det}$ showing strong correlations with their simulated values in both horizon (H) conditions (For $\sigma_{ran}$, $R = 0.91 (\mbox{H} = 1) \mbox{ and } 0.84 (\mbox{H} = 6), p < 0.001$, For $\sigma_{det}$, $R = 0.60 (\mbox{H} = 1) \mbox{ and } 0.59 (\mbox{H} = 6), p < 0.001$). However, while the relationship was near perfect for random noise ($\frac{\mbox{\small Recovered } \sigma_{ran}}{\mbox{\small Simulated } \sigma_{ran}} = 1.01$), there was a systematic bias to underestimate the level of deterministic noise by about 32\% ($\frac{\mbox{\small Recovered } \sigma_{det}}{\mbox{\small Simulated } \sigma_{det}} = 0.68$). Despite this underestimation of deterministic noise in both horizon conditions, the difference in deterministic noise between horizons is much better captured (see Supplementary Materials section 2.2). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizon conditions. In addition, we see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. 		
	% $\frac{\mbox{\small Recovered } \Delta \sigma_{det}}{\mbox{\small Simulated } \Delta \sigma_{det}} = 0.74$

In addition to the conventional subject-level parameter recovery analysis presented here, we also performed parameter recovery analysis that examined how faithful the full posterior distribution of group-level parameters can be recovered in simulated data (Supplementary Figure S5, S6, S9, S10). Qualitatively, we also showed that our way of modeling deterministic noise is capable of capturing known deterministic processes intentionally omitted from the full model (Supplementary Figure S4). Full details of these additional analysis are presented in the Supplementary Materials.

@@

\subsubsection*{Model-based results} 
Posterior distributions over the group-level means of the deterministic and random noise standard deviation $\sigma_{det}$ and $\sigma_{ran}$ are shown in Figure \ref{fig:mb1} and Supplementary Figure S11. Consistent with our model-free results, we see that both random and deterministic noise are non-zero. Numerically, random noise is about 2-3 times larger than the deterministic noise. By computing the posterior distribution of $\sigma^2_{det}/(\sigma^2_{det}+\sigma^2_{ran})$, our data suggests that 14.25\% of the variability in random exploration is accounted for by deterministic noise ([4.90\%, 28.81\%], 95\% CI). In addition, we find that both random and deterministic noise increase with horizon. This increase was larger for random noise (mean = 7.13, 100\% of samples showed an increase in random noise with horizon) than deterministic noise (mean = 2.59, 98.64\% of samples showed an increase in deterministic noise with horizon). But intriguingly, the relative increase in both types of noise was similar (Figure \ref{fig:ratio}). That is, when we compute the relative increase in deterministic noise with horizon, $\sigma^{det}_{horizon6}/\sigma^{det}_{horizon1}$, it is very similar to the relative increase in random noise with horizon $\sigma^{ran}_{horizon6}/\sigma^{ran}_{horizon1}$.

\added[id=siyu]{Similar results are found in other variants of our model. Results for a model that estimates random and deterministic noises separately for [1 3] and [2 2] conditions, and a model with an alternative definition of information bonus $dI$, can be found in the Supplemental Materials (Supplemental Figure S12, S13). }

\added[id=siyu]{To ensure that the joint increase of random and deterministic noises is genuine and not an artifact from the fitting procedure, we computed the correlation between ground-truth values of random noise, and best-fitting values of deterministic noise (and vice versa), and they do not correlate (Supplementary Figure S14). Furthermore, we simulated data from a series of reduced models with known random and deterministic noise values in which either random or deterministic noise does not change with Horizon, and fit our model to the simulated data. Our model detects and only detects a change in random/deterministic noise with horizon, when the change is present in the model that simulates the data (Supplementary Figure S15, S16).}

@@

As a control, we also applied posterior predictive checks on alternative models that consider only deterministic or only random noise, and these reduced models fail to capture all qualitative patterns (Supplementary Figure S17, S18). Full details of this analysis can be found in Supplementary Materials section 3.5.

@@

Although the random noise interpretation is theoretically appealing, our approach, while an improvement on previous methods, is not without limitations. Most important is that our measure of `random' noise is only an upper bound on the true level of randomness and that, in principle, the random decision noise could be lower. Specifically, in our model, what we labeled random noise was really `non-stimulus-driven variability'. While this non-stimulus-driven variability could be driven by truly random stochastic processes, it could also be driven by deterministic processing that is unrelated to the stimuli in the task. For example, such deterministic noise could be driven by differences in where people look, or for how long they look, or by whether they were fidgeting or scratching their nose \citep{Musall2019}. \added[id=siyu]{Another limitation is that deterministic noise is defined based on the stimulus within a game, between-game deterministic strategies and stimuli from previous games (e.g., memory of previously seen game) were also treated as random noises in our model, although our model could be potentially extended by considering deterministic noise over a sequence of stimuli across games \citep{WYART2016, Wyart_2018}.} In addition to this conceptual limitation in measuring deterministic noise, parameter recovery simulations suggest that our estimation method also slightly underestimates deterministic noise (see Figure \ref{fig:paramrecover_main}, Supplementary Figure S5, S6). As a result, from both a conceptual and methodological perspective, it is possible that the remaining 86\% of the decision noise that is not stimulus-driven noise, could be deterministic.

@@

\added[id=siyu]{We show first that a change in noise is mathematically equivalent to a change in reward signal strength in our decision model (see also \cite{Cinotti19}).}. To show how a change in reward processing could affect random and deterministic noise, consider the simple decision model we introduced in Equation \ref{eq:newmodel}. In this model, choice is determined by the sign of the difference in utility $\Delta Q$ between the two options, where
\begin{equation}
	\label{eq:noise1}
	\Delta Q = 
	\Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Now imagine a case where the reward signal is scaled by a factor $\beta$.  In this case, $\Delta Q$ becomes
\begin{equation}
	\label{eq:noise2}
	\Delta Q = 
	\beta \Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Because the choice only depends on the sign of $\Delta Q$, scaling $\Delta Q$ by a factor of $1/\beta$ will not change the behavior of the model.  Thus, if we divide both sides of the above equation by $\beta$ we get
\begin{equation}
	\label{eq:noise3}
	\Delta Q / \beta = 
	\Delta R + A \Delta I / \beta + b / \beta + n_{det} / \beta + n_{ran} / \beta
\end{equation}
which is equivalent to a scaling of both deterministic and random noise by the same factor $1/\beta$. Thus, one interpretation of our result that both deterministic and random noise change across horizons with the same ratio, is that this reflects a change in reward processing. That is, the reward signal is reduced in the longer horizon condition (smaller $\beta$ in horizon 6 than horizon 1).  %In the alternate interpretation, where deterministic and random noise are varied independently it is not clear why they should change by the same scale factor.

### 2025-04-21 Plos Comp Bio - revision v1/supplemental/DetRanNoise_Supplementary.tex
u

	\subsection{Model-free analysis with simulated choice from a pure random noise model}
	\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RDBayes_pinconsistent_validation.jpg}
		\caption{Model-free analysis with simulated choices from a model that has only random noise validates our prediction of p(inconsistent) for pure random noise. The extent to which simulated choices are inconsistent completely overlaps with our pure random noise prediction(p $>$ 0.05). This suggests that when choice inconsistency lies below the pure random noise prediction indeed provides evidence that deterministic noise exists in random exploration (Figure 3).}
		\label{fig:s3}
	\end{center}
	\end{figure}
	\newpage
	\section{Model validation analyses}
	\subsection{Deterministic noise captures known deterministic processes excluded from the model}	
	We checked if our fitted deterministic noise could indeed capture unobserved deterministic process that was not accounted for by the decision model. We test this by leaving out one known deterministic process from the decision model, and ask if our method could recover that known deterministic process as deterministic noise. In particular, we fit a reduced version of our model that only considers reward and ignores the influence of uncertainty condition and spatial bias on explore-exploit decisions.  
	$$\Delta Q= \Delta R+n_{det}+n_{ran}$$
	Here, $\Delta Q, \Delta R, n_{det}, n_{ran}$ represent the same variables as in the full model (Equation 2 in the manuscript).	If deterministic noise in our model can indeed capture unobserved deterministic processes that's missed by the model, then we would expect to see a higher level of fitted deterministic noise in the reduced model compared to in the full model, whereas the level of random noise should remain unchanged. By comparing the	fitted posterior distributions over the group-level means of the deterministic and random noise parameters $\sigma_{det}$ and $\sigma_{ran}$, as expected, we observed an increase in deterministic noise and no change in random noise between the reduced and the full model (Figure \ref{fig:reducedmodel}). This suggests that our model is capable of detecting missing deterministic processes.

	In general, increasing random noise leads to higher p(low mean) and lower p(consistent) between repeated games, and increasing deterministic noise leads to higher p(low mean) and higher p(consistent).

@@

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=0.65\textwidth]{figures/R1_shuffle_shuffled_recovery.jpg}
			\caption{Parameter recovery for shuffled data. To show that the joint increase of random and deterministic sources of noise is not caused by a limitation of the fitting procedure, we calculated the correlation between ground-truth values of random noise, and best-fitting values of deterministic noise (and vice versa). Ground-truth values are shuffled best-fit parameters. As expected, ground-truth random values do not correlate with recovered deterministic noises, showing that the increase of deterministic noise with horizon is genuine and not a by-product of increase of random noise, and vice versa.  }
			\label{fig:paramrecover_sub}
		\end{center}
	\end{figure}

	\newpage	
	\subsection{Parameter recovery over a bigger range of arbitrary combinations of deterministic and random noises \label{gridrecovery}}

@@

	\section{Additional model-based analyses}
	\subsection{Replication of Figure 5 without excluding subjects}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/RDBayes_hyperprior__all.jpg}
			\caption{Model based analysis with data from all participants (i.e. no exclusions) showing the posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and change with horizon (B, D).  However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.}
			\label{fig:s8}
		\end{center}
	\end{figure}
	\newpage
	\subsection{Alternative models: variations of the two-noise model}
	To check whether all aspects of the model were necessary to reproduce the qualitative pattern of findings, we also built and fit five additional versions of the model. These models varied in whether deterministic and random noise are present or not and whether either types of noise is dependent on horizon. Specifically, we tested the following 6 models (Note that the $\sigma^{ran}_{horizon},\sigma^{det}_{horizon}$ model is our original full model).

@@

	We again simulated choices using fitted parameters from these models and repeated the model-free analysis on the simulated data. As shown in Figure \ref{fig:posteriorcheck6models}, only one of these alternative models, where random noise is horizon dependent but deterministic noise is not, can capture the full qualitative pattern of behavior. However, the quantitative fit to the data is not as good (Figure \ref{fig:posteriorcheck6models}).% SIYU says: they actually look quite similar, without horizon dependent deterministic noise, the fixed det noise model is only marginally worse on p(inconsistent) in [2 2] condition, as it overlaps with the pure random noise line more.

	In our model, the subject-level noises are assumed to follow a gamma distribution (to ensure positiveness), the posteriors are right skewed (maximal likelihood estimation or mode is smaller than the mean), Because of the skewness, simulating data with the “mean” is nosier than simulating with the true distribution. Simulating from the true distribution requires taking expected value over all possible noise values, for ease of implementation, we simulated data by taking random samples from the posterior distribution (instead of using the mean). The simulation was repeated 50 times and then averaged (Supplementary Figure \ref{fig:posteriorcheck6models}). Simulation using the "mode" or maximal likelihood estimation was also shown here (Supplementary Figure \ref{fig:posteriorcheck6models2}).

@@

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=0.8\textwidth]{figures/R1_2noise_6modelcomparison_dist.jpg}
			\caption{
				Model comparison. A-D. both deterministic and random noise are horizon dependent, E-H. only random noise is horizon dependent, I-L. only deterministic noise is horizon dependent, M-P. neither random nor deterministic noise is horizon dependent, Q-T. only deterministic noise is assumed to be present, U-X. only random noise is assumed to be present.}
			\label{fig:posteriorcheck6models}
		\end{center}
	\end{figure}
	\begin{figure}[hp]
	\begin{center}
		\includegraphics[width=0.8\textwidth]{figures/R1_2noise_6modelcomparison_map.jpg}
		\caption{
			Model comparison (using maximal likelihood estimation). A-D. both deterministic and random noise are horizon dependent, E-H. only random noise is horizon dependent, I-L. only deterministic noise is horizon dependent, M-P. neither random nor deterministic noise is horizon dependent, Q-T. only deterministic noise is assumed to be present, U-X. only random noise is assumed to be present.}
		\label{fig:posteriorcheck6models2}
	\end{center}
	\end{figure}

	Moreover, we examined if our model can indeed qualitatively capture whether deterministic and random noise are present or not and whether either types of noise is dependent on horizon. To test this, we simulated choices from each of the 6 models, and then fit the simulated choices with our original full model. The simulation was repeated 50 times for each model. Indeed, we showed that our model can capture both the existence of random and deterministic noise, and whether each noise changes with horizon condition (Figure \ref{fig:6model}), with only one exception that our model falsely detected a small fraction of deterministic noise when no deterministic noise was present (Figure \ref{fig:6model}). This phenomenon was also examined and discussed in the section \ref{gridrecovery} above.

@@

	\newpage
	\subsection{Alternative models: separate noises for [1 3] and [2 2] condition}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/R1_hyperprior.jpg}
			\caption{Model based analysis from a model that estimates random and deterministic noises separately for [1 3] and [2 2] conditions. The posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, E) and deterministic (C,G) noises are nonzero (A, C, E, G) and change with horizon (B, D, F, H).  However, random noise has both a greater magnitude overall (A, E) and a greater change with horizon (B, F) than deterministic noise. Both random and deterministic noises are higher in [1 3] compared to [2 2] conditions.}
			\label{fig:s8}
		\end{center}
	\end{figure}
	\newpage
	\subsection{Alternative models: use variance differences as dI}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/R1_hyperprior_dIvar.jpg}
			\caption{Model based analysis from a model that uses variance differences as dI. The posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and change with horizon (B, D).  However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.}
			\label{fig:s8}
		\end{center}
	\end{figure}

>>>>

	\subsection{Control analysis: model-free analysis with simulated choice from a random-noise only model}
	\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RDBayes_pinconsistent_validation.jpg}
		\caption{Model-free analysis with simulated choices from a model that has only random noise validates our prediction of p(inconsistent) for pure random noise. The extent to which simulated choices are inconsistent completely overlaps with our pure random noise prediction(p $>$ 0.05). This suggests that when choice inconsistency lies below the pure random noise prediction indeed provides evidence that deterministic noise exists in random exploration (Figure 3).}
		\label{fig:s3}
	\end{center}
	\end{figure}
	\newpage
	\section{Model validation analyses}
	\subsection{Deterministic noise captures known deterministic processes excluded from the model}	
	We checked if our fitted deterministic noise could indeed capture unobserved deterministic process that was not accounted for by the decision model. We test this by leaving out one known deterministic process from the decision model, and ask if our method could recover that known deterministic process as deterministic noise. In particular, we fit a reduced version of our model that only considers reward and ignores the influence of uncertainty condition and spatial bias on explore-exploit decisions.  
	$$\Delta Q= \Delta R+n_{det}+n_{ran}$$
	Here, $\Delta Q, \Delta R, n_{det}, n_{ran}$ represent the same variables as in the full model (Equation 2 in the manuscript).	If deterministic noise in our model can indeed capture unobserved deterministic processes that's missed by the model, then we would expect to see a higher level of fitted deterministic noise in the reduced model compared to in the full model, whereas the level of random noise should remain unchanged. By comparing the	fitted posterior distributions over the group-level means of the deterministic and random noise parameters $\sigma_{det}$ and $\sigma_{ran}$, as expected, we observed an increase in deterministic noise and no change in random noise between the reduced and the full model (Figure \ref{fig:reducedmodel}). This suggests that our model is capable of detecting missing deterministic processes.

@@

	\newpage	
	\subsection{Parameter recovery of arbitrary combinations of deterministic and random noises \label{gridrecovery}}

@@

	\section{Additional model-based analyses}
	\subsection{Replication of Figure 5 without excluding subjects}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/RDBayes_hyperprior__all.jpg}
			\caption{Model based analysis with data from all participants (i.e. no exclusions) showing the posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and change with horizon (B, D).  However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.}
			\label{fig:s8}
		\end{center}
	\end{figure}
	\newpage
	\subsection{Alternative model: fit separate random and deterministic noises for [1 3] and [2 2] conditions}

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/R1_hyperprior.jpg}
			\caption{Model based analysis from a model that estimates random and deterministic noises separately for [1 3] and [2 2] conditions. The posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, E) and deterministic (C,G) noises are nonzero (A, C, E, G) and change with horizon (B, D, F, H).  However, random noise has both a greater magnitude overall (A, E) and a greater change with horizon (B, F) than deterministic noise. Moreover, both random and deterministic noises have a greater magnitude in [1 3] compared to [2 2] conditions.}
			\label{fig:s8}
		\end{center}
	\end{figure}
	\newpage
	\subsection{Alternative model: use variance differences as dI}
	Here, we fit a model with an alternative definition of $\Delta I$ that $\Delta I$ is defined to be difference between the variances of rewards shown on the forced-choice trials.
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/R1_hyperprior_dIvar.jpg}
			\caption{Model based analysis from a model that uses variance differences as dI. The posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and change with horizon (B, D).  However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.}
			\label{fig:s8}
		\end{center}
	\end{figure}
	\newpage

	\subsection{Examine the co-variance of random vs deterministic noise}
	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=0.65\textwidth]{figures/R1_shuffle_shuffled_recovery.jpg}
			\caption{Parameter recovery for shuffled data. To show that the joint increase of random and deterministic sources of noise is not caused by a limitation of the fitting procedure, we calculated the correlation between ground-truth values of random noise, and best-fitting values of deterministic noise (and vice versa). Ground-truth values are shuffled best-fit parameters. As expected, ground-truth random values do not correlate with recovered deterministic noises, showing that the increase of deterministic noise with horizon is genuine and not a by-product of increase of random noise, and vice versa.  }
			\label{fig:paramrecover_sub}
		\end{center}
	\end{figure} 
	\newpage

	\subsection{Alternative models: variations of the two-noise model}	
	In this section, we examined 6 model variants that varied in whether deterministic and random noise are present or not and whether either types of noise is dependent on horizon. Specifically, we tested the following 6 models (Note that the $\sigma^{ran}_{horizon},\sigma^{det}_{horizon}$ model is our original full model).

@@

@@

	We then examined if our model can indeed qualitatively capture whether deterministic and random noise are present or not and whether either types of noise is dependent on horizon. To test this, we simulated choices from each of the 6 models, and then fit the simulated choices with our original full model. The simulation was repeated 50 times for each model. Indeed, we showed that our model can capture both the existence of random and deterministic noise, and whether each noise changes with horizon condition (Figure \ref{fig:6model}), with only one exception that our model falsely detected a small fraction of deterministic noise when no deterministic noise was present (Figure \ref{fig:6model}). This phenomenon was also examined and discussed in the section \ref{gridrecovery} above.

@@

	%we also built and fit five additional versions of the model.

	To check whether all aspects of the model were necessary to reproduce the qualitative pattern of findings, we again simulated choices using fitted parameters from these models and repeated the model-free analysis on the simulated data. As shown in Figure \ref{fig:posteriorcheck6models}, only one of these alternative models, where random noise is horizon dependent but deterministic noise is not, can capture most of the qualitative features, including the increase in p(high info), p(low mean) and p(consistent). However, the quantitative fit to the data is not as good (Figure \ref{fig:posteriorcheck6models}). Moreover, when comparing p(consistent) with the theoretically predicted p(consistent) for pure random noise (see analysis in Figure 3), only the fully model is statistically significant in both horizon conditions. Out of 50 simulations, the best alternative model only shows statistical significance in half of the simulations in Horizon 6, and fails on the other 50\% of the times. Because deterministic noise is not allowed to change with horizon in this alternative model, in horizon 6, deterministic noise is not large enough compared to the magnitude of random noise, to make p(inconsistent) lower than the random-noise-only prediction.   % SIYU says: they actually look quite similar, without horizon dependent deterministic noise, the fixed det noise model is only marginally worse on p(inconsistent) in [2 2] condition, as it overlaps with the pure random noise line more.

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=0.8\textwidth]{figures/R1_2noise_6modelcomparison_dist.jpg}
			\caption{
				Model comparison. A-D. both deterministic and random noise are horizon dependent, E-H. only random noise is horizon dependent, I-L. only deterministic noise is horizon dependent, M-P. neither random nor deterministic noise is horizon dependent, Q-T. only deterministic noise is assumed to be present, U-X. only random noise is assumed to be present.}
			\label{fig:posteriorcheck6models}
		\end{center}
	\end{figure}
	\newpage

	In our model, the subject-level noises are assumed to follow a gamma distribution (to ensure positiveness), the posteriors are right skewed (maximal likelihood estimation or mode is smaller than the mean), Because of the skewness, simulating data with the “mean” is noisier than simulating with the true distribution. Simulating from the true distribution requires taking expected value over all possible noise values, for ease of implementation, we simulated data by taking random samples from the posterior distribution (instead of using the mean). The simulation was repeated 50 times and then averaged (Supplementary Figure \ref{fig:posteriorcheck6models}). Simulation using the "mode" or maximal likelihood estimation was also shown here (Supplementary Figure \ref{fig:posteriorcheck6models2}) and produces closer fit to the data.

	\begin{figure}[hp]
	\begin{center}
		\includegraphics[width=0.8\textwidth]{figures/R1_2noise_6modelcomparison_map.jpg}
		\caption{
			Model comparison (using maximal likelihood estimation). A-D. both deterministic and random noise are horizon dependent, E-H. only random noise is horizon dependent, I-L. only deterministic noise is horizon dependent, M-P. neither random nor deterministic noise is horizon dependent, Q-T. only deterministic noise is assumed to be present, U-X. only random noise is assumed to be present.}
		\label{fig:posteriorcheck6models2}
	\end{center}
	\end{figure}
